% !TeX spellcheck = en_GB
%% 
%% Copyright 2007-2018 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%% 
%%
%% $Id: elsarticle-template-num.tex 64 2013-05-15 12:23:51Z rishi $
%%
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{mathpartir,color}
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{graphicx}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\symb}[1]{\textit{#1}} 
\newcommand{\noop}{\symb{Noop}}
\newcommand{\Push}{\symb{Push}}
\newcommand{\Pull}{\symb{Pull}}
\newcommand{\while}{\symb{While}}
\newcommand{\cond}{\symb{cond}}
\DeclareMathOperator{\vars}{vars}
\newcommand{\isvalid}{\symb{isValid}}
\newcommand{\isremvalid}{\symb{remIsValid}}
\newcommand{\rem}[1]{\symb{rem}(#1)}
\newcommand{\IF}[3]{\symb{if}\,(#1)~#2~\symb{else}~#3 }
\newcommand{\feval}[2]{\llbracket#1\rrbracket_{#2}}
\newcommand{\True}{{\tt True}}						
\newcommand{\False}{{\tt False}}				
\newcommand{\transl}[1]{\llbracket#1\rrbracket}
\newtheorem{definition}{Definition}
\newtheorem{Property}{Property}
\newtheorem{Theorem}{Theorem}
\newcommand{\abs}[1]{#1^\#}
\newcommand{\AM}{\mathcal{M}}
\newcommand{\Prog}{\mathcal{P}}
\usepackage{etoolbox}
\DeclareMathOperator{\range}{range}
\DeclareMathOperator{\dom}{dom}
\newcommand{\Overlap}[1]{O(#1)}

%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

\journal{Journal of Logical and Algebraic Methods in Programming}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}
\title{Leveraging Access Mode 
Declarations in a Model for Memory Consistency in Heterogeneous Systems }
%\title{Ensuring Memory Consistency in Heterogeneous Systems Based on Access Mode
%Declarations }


%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}
\author[i3s]{Ludovic Henrio\corref{cor1}}
\ead{ludovic.henrio@cnrs.fr}
\author[liu]{Christoph Kessler}
\ead{christoph.kessler@liu.se}
\author[liu]{Lu Li}
\ead{lu.li@liu.se}
\cortext[cor1]{Corresponding author}

\address[i3s]{Universit\'e~C\^ote~d'Azur, CNRS, I3S, France.}
\address[liu]{University of Linköping, Sweden}

\begin{abstract}
Running a program on disjoint memory spaces requires to address memory consistency 
issues and to perform  transfers so that the program always accesses  the right 
data. Several approaches exist to ensure the consistency of the memory accessed,  we 
are interested here in the verification of a declarative approach where each component of 
a computation is annotated with an access mode declaring which part of the memory is 
read or written by the component. The programming framework uses the component
annotations to guarantee the validity of the memory accesses. This is the mechanism used in VectorPU, a C++ library for programming CPU-GPU heterogeneous  systems 
and this article proves the correctness of the software cache-coherence mechanism used in 
the library.  The formalism we propose also takes into account arrays for which a single validity status is stored for the whole array; additional mechanisms for dealing with overlapping arrays are also studied. Beyond the scope of VectorPU, this article can be considered as a simple 
and effective formalisation of memory consistency mechanisms based on the explicit 
declaration of the effect of each component on each memory space.

\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Memory consistency \sep CPU-GPU heterogeneous systems \sep  
data transfer \sep  software caching \sep  cache coherence 
%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

\section{Introduction}

\input{vectorpu-intro.tex}

\section{A Formalization for Reasoning on  Consistency in VectorPU}\label{sec:Formal}

In this section we provide a minimal calculus to reason on the memory operations that can 
exist in a framework that deals with memory consistency like VectorPU. We first define a 
set of effects that operations can have on the consistency of the memory. Then we define a 
small calculus expressing different memory accesses and their composition into complex 
procedures. Finally, we express VectorPU operations as higher-level statements that can 
be translated  into the core calculus, and show that if all memory 
accesses are annotated correctly through VectorPU annotations the program cannot try to 
access an invalid data and the memory spaces are put in coherence when needed. We also 
show that VectorPU tracks the validity status of the memory adequately. In this 
section we abstract away the values stored in memory and we 
do not deal with any form of aliasing. A more precise analysis of effects and aliasing is 
out of the scope of this paper, it could be for example inspired 
from~\cite{Nielson1999}.
We place ourself in a simplified setting where each variable is hosted in exactly two 
memory locations, e.g.\ a CPU (main) memory and a GPU memory location, but the work could be extended to multiple memory 
locations without any major difficulty.

\subsection{An effect system for consistency between memory locations}
We start from a simple effect system, it expresses the effect of writing or reading a 
memory 
location on the consistency status of the memory. Each location is either in  
\textit{valid} state when it holds a usable data 
or \textit{invalid} state when the value at the location is not valid anymore.

We express five  operations: reading, writing, 
\Push\ for uploading the local memory location into the other one, and \Pull\ for the 
contrary. \noop\ is an operation that does nothing.
 \[E::= \Push ~|~ \Pull ~|~ r~|~ w ~|~ %rw ~|~ 
\noop\]

The effect of these operations express their requirements and effects on a single memory 
location. We 
express 
below the semantics of each of the operations on the consistency status of the concerned 
memory location. 
%The status memory location 
The \textit{memory status of a variable}
is a pair of the status of its locations, 
where each status is 
either $V$ for valid or $I$ for invalid. The first element is the status of the local 
memory, and the second one is the status of the remote memory. For example, for a 
program running on a CPU while the remote memory is a GPU, a status $(V,I)$ means that 
the memory is valid and can be read on the CPU, but is invalid on the GPU and should be 
transferred before being usable there.

Each operation has a signature in the sense that it may require a certain memory status 
and 
will produce another memory status. The signature of each operation is expressed below.
We use
variables $X$, $Y$, $Z$, $T$ that are considered as universally quantified in each rule. 
They can 
be instantiated with either $V$ or $I$.
\begin{mathpar}
\Push: (V,X)\mapsto (V,V)

\Pull: (X,V)\mapsto (V,V)

r: (V,X)\mapsto (V,X)

%rw: (V,X)\mapsto (V,I)

w: (X,Y)\mapsto (V,I)

\noop: (X,Y)\mapsto (X,Y)
    \end{mathpar}

These signatures are effects  expressing that
$r$ is a reading operation requiring validity of data and ensuring not to modify it, 
the distant status is unchanged; $w$ 
is a  writing operation that modifies data locally but do not require validity, it 
invalidates the remote memory.  \Push\ uploads the local memory and thus makes valid the 
distant memory; 
it requires that the data is locally valid, and \Pull\ is the symmetrical operation.

An additional operation could be defined: 
 an 
$rw$ operation would represent a read and/or write access, it would both require data 
validity and invalidate 
remote status: $(V,X)\mapsto (V,I)$. This operation is however not needed here but we will have a similar one at the annotation level, see below.

\subsection{A language for modelling consistency and effects}\label{sec-core}
We now create a core calculus to be able to reason on programs involving sequences of 
effects on different memory locations. $x,y$ range over variables and we introduce  statements manipulating 
variables. We use sequence and simple loops and conditionals. 
Operations with effects 
now apply to a variable;  the $\rem{E~x}$  is a remote operation 
 on the remote memory. For example, a GPU procedure 
writing $x$ and reading $y$ would correspond to the pseudo-code: $\rem{w~x};\rem{r~y}$. 
Statements $S$ are defined as:

{\small \[S::=E~x~|~\rem{E~x}~|~S;S'~|~\while(\cond) S ~|~ \IF{\cond}{S}{S'}\]}
\noindent where $E$ $x$ denotes some effect $E$ on variable $x$, with $E\in \{r,$ $w,$  
$\Push,$ $\Pull,$ $\noop\}$.

We are  interested  in conditionals dealing 
with the validity status of the variables. Other conditionals  are expressed as a generic 
binary operator $\oplus$ but  operators with different arities could be added as 
well:
\[\cond::=\isvalid~x~|~\isremvalid~x~|~x\oplus y\]
%$\oplus$ replaces operations on values; we are not interested in evaluating them ; only 
%to show that while can use values (perhaps to be improved)
\noindent where \textit{isValid} $x$ and \textit{remIsValid} $x$ denote checks of the validity status flag of the local and remote location of $x$, respectively.

We now define a small step operational semantics for our core calculus.
It relies on the validity status of variables, recorded in a store $\sigma$ mapping 
variable names to validity pairs. Semantics is written as a transition relation between 
pairs consisting of a statement and a store: $(S,\sigma)$. The sequencing operator $;$ is associative with \noop\ as a neutral element. 
Consequently each non-empty sequence of instruction can be rewritten as $S;S'$ where $S$ 
is neither a 
sequence nor \noop. $\sigma[x\mapsto (X,Y)]$ is the update operation on maps. 
%The swap 
%operation on 
%pairs is extended to store (it swaps the value associated to each variable)


\begin{figure*}[tp]
\begin{mathpar}
\inferrule[valid]
{\sigma(x)=(V,X)}
{\feval{\isvalid~x}{\sigma}=\True}

\inferrule[invalid]
{\sigma(x)=(I,X)}
{\feval{\isvalid~x}{\sigma}=\False}

\inferrule[rem-valid]
{\sigma(x)=(X,V)}
{\feval{\isremvalid~x}{\sigma}=\True}

\inferrule[rem-invalid]
{\sigma(x)=(X,I)}
{\feval{\isremvalid~x}{\sigma}=\False}

\inferrule[Effect]
{\sigma(x)=(X,Y) \\ E:  (X,Y)\mapsto (Z,T)  }
{(E~x;S,\sigma)\to (S,\sigma[x\mapsto(Z,T)])}


\inferrule[Remote Effect]
{\sigma(x)=(X,Y) \\ E:  (Y,X)\mapsto (Z,T) }
{(\rem{E~x};S,\sigma)\to (S,\sigma[x\mapsto(T,Z)])}

\inferrule[While-True]
{\feval{\cond}{\sigma} }
{(\while(\cond) S;S',\sigma)\to (S;\while(\cond) S;S',\sigma)}

\inferrule[While-False]
{\neg\feval{\cond}{\sigma} }
{(\while(\cond) S;S',\sigma)\to (S',\sigma)}

\\

\inferrule[IF-True]
{\feval{\cond}{\sigma} }
{((\IF\cond S S');S'',\sigma)\to (S;S'',\sigma)}

\inferrule[IF-False]
{\neg\feval{\cond}{\sigma} }
{((\IF\cond S S');S'',\sigma)\to (S';S'',\sigma)}

    \end{mathpar}
\caption{Operational semantics of validity status.}\label{fig:Opsem}
\end{figure*}

The semantics is presented in Figure~\ref{fig:Opsem}. Like in the previous section, we 
use validity variables $X$, $Y$, $Z$, $T$ 
that are universally quantified in each rule.
 The first %three
 four rules present the 
evaluation of conditional statements, we suppose additional rules exist for evaluating 
$\oplus$\footnote{We are only interested in cache consistency properties, we thus suppose  that evaluation of $\oplus$ always succeed, and in particular  variables accessed by the operation are specified as a $r$ operation preceding the condition.}. The next rule applies an effect on a variable $x$ updating the validity store, 
and the \textsc{Remote Effect} rule applies an effect occurring on the distant memory, it 
applies the symmetric of the effect to the variable. Note that \Push\ is the symmetric of 
\Pull\ and we could have removed one of those two operations without loss of generality. 
The last rules are standard ones for \symb{if} and \symb{while} statements.

\noindent\emph{Initial state:} To evaluate a sequence of statements $S$ using the 
variables 
$\vars(S)$, we put it in a configuration with 
an initial 
store where data is hosted on the CPU and all variables are initially mapped to $(V,I)$: 
$\sigma_0=(x\mapsto 
(V,I))^{x\in \vars(S)}$.


%for the moment 2 address spaces, see if we need many of them

A configuration is \emph{reachable} if it is possible to obtain this configuration 
starting from the initial configuration and applying any number of reductions: 
$(S,\sigma)$ is reachable if $(S,\sigma_0)\to^*(S',\sigma)$ where $\to^*$ is the 
reflexive 
transitive closure of $\to$. We write $(S,\sigma)\not\to $ and say that the configuration 
is \emph{stuck} if no reduction rule can be 
applied on $(S,\sigma)$.
\begin{Property}[Progress]\label{prop:stuck}
 A configuration is stuck if the validity status of the accessed variable is 
incompatible with the effect to be applied\footnote{We say that there is no unification  between $X$ and $Y$ if one of the two variables must have the value $V$, and the other one the value $I$. This relation is extended to pairs of variables.}:
\[\begin{array}{@{}l@{}}
(S,\sigma)\not\to\ \iff~
 \begin{array}[t]{@{}l@{}}
S=E~x;S' \land \sigma(x)=(X,Y) 
							\land E:  (X',Y')\mapsto (Z,T)\ 
						\\\qquad 	\land~\text{there is no unification  between } 
							(X,Y)   
							\text{ and } (X',Y')\\
\lor S\!=\!\rem{E~x};S' \land \sigma(x)\!=\!(X,Y) 
							\land E:  (X',Y')\!\mapsto\! (Z,T)\ 
							\\\qquad \land~\text{there is no unification  between } 
							(X,Y)   
							\text{ and } (Y',X')		
\end{array}
\end{array}\]
Note that this supposes that $\oplus$ always succeeds.
\end{Property}
\begin{proof}[Proof sketch]
By case analysis on the first statement of $S$, there is always 
one rule applicable provided the premises of the rule can be evaluated.
In the case of the last four rules this requires the evaluation of \cond. If $\oplus$ 
always succeeds then \cond\ can always be 
evaluated. The only case remaining is if there is no unification possible between the 
effect of an operation and the current validity status of the affected variable, this 
concerns the rule \textsc{Effect} and \textsc{Remote Effect} and corresponds to the two 
cases expressed in the theorem.
\end{proof}


\begin{Property}[Safety]\label{prop:safe}
A state is said to be \emph{unsafe} if at least one variable is mapped to 
$(I,I)$.
It is impossible to reach an unsafe state from the initial state.
\end{Property}
\begin{proof}[Proof sketch] 
Unsafe states are avoided  because of the effects of operations: 
only effect rules modify the store and no effect can reach $(I,I)$, except 
\noop\ starting from $(I,I)$, whic is sufficient to conclude as the inital state is not $(I,I)$.
\end{proof}

\noindent
   \emph{Example:} 
    $w x;\rem{r~x}$ cannot be fully evaluated. Indeed, $(w x;\rem{r~x},(x\mapsto 
    (V,I)))\to (\rem{r~x},(x\mapsto (V,I)))$, but  $\rem{r~x}$ requires that $x$ is 
    mapped to $(X,V)$ for some $X$ which is not the case.
However if we add a \Push\ operation to ensure the validity of the accessed memory the 
program $w~x;Push;\rem{r~x}$ can be reduced as follows:\\
$(w x;\rem{r~x},(x\mapsto (V,I)))\\~\qquad\to (\Push;\rem{r~x},(x\mapsto 
(V,I)))\\~\qquad\to 
(\rem{r~x},(x\mapsto (V,V)))\to (\noop,(x\mapsto (V,V)))$

%\subsection{Protecting Memory Accesses}
%We now define protected operations as macros that can be compiled into the core 
%calculus. 
%The objective of this section is to show that if each memory access is protected, i.e.\ 
%if 
%a program only contains protected effects, and no direct memory effect, then the program 
%executes safely without getting stuck in a configuration. We propose an extended 
%language 
%with the following syntax:
%\[S'=S~|~R~x~|~W~x~|~RW~x|~\rem {R~x}~|~\rem {W~x}~|~\rem {RW~x}\]
%For each memory effect, we now have a protected access; for example, $R~x$ denotes a 
%protected reading operation that ensures that the validity status is correct before 
%performing the read. We have protected local accesses and protected accesses on the 
%distant memory. Similarly to the VectorPU library, the protected accesses can be 
%considered as macros and the programs of the extended syntax can be translated into the 
%core syntax as follows:
%\begin{mathpar}
%\transl{R~x}=(\IF{\isvalid~x}{\noop}{\Pull~x});r~x
%
%\transl{\rem {R~x}}=(\IF{\isremvalid~x}{\noop}{\Push~x});\rem{r~x}
%
%\transl{RW~x}=(\IF{\isvalid~x}{\noop}{\Pull~x});rw~x
%
%\transl{\rem{RW~x}}=(\IF{\isremvalid~x}{\noop}{\Push~x});\rem{rw~x}
%
%\transl{W~x}= w~x
%
%\transl{\rem{W~x}}= \rem{w~x}
%\end{mathpar}
%This encoding corresponds  to the macros instantiated in VectorPU except that VectorPU 
%additionally tracks the effects, and this is unnecessary here because our semantics is 
%tracking the effect. Note that it is easy to check that VectorPU tracks the effects in 
%the same way as our semantics does. These translation rules  perform \Push\ or 
%\Pull\ operations in order to ensure that the memory is in a correct validity status for 
%the read or write operation to be performed.
%
%\begin{Theorem}[Protecting operations ensures progress]
%If a program $S'$ is written with only  protected operations, i.e. no $r$, $w$, $rw$, 
%$\Push$ or $\Pull$, then its execution cannot reach a stuck configuration.
%\end{Theorem}
%
%\begin{proof}[Proof sketch] By Property~\ref{prop:stuck}, it is sufficient to prove that 
%unification on the validity status is always possible, in other words, if a program has 
%only protected operations then this unification is always possible. 
%
%This is done by case analysis of the effect applied in the translated program, and 
%showing that in each case the effect rule can be applied.
%Take the example of \Pull. \Pull\ only appears upon $R~x$ or $RW~x$ and when  
%$\isvalid~x$ is 
%false. In this case $\sigma(x)=(I,V)$ by property~\ref{prop:safe}. Consequently \Pull\ 
%is 
%not stuck.
%Now considering the effect $r$, $r~x$ can only be obtained by translation of $R~x$. 
%Consequently, $r~x$ appears necessarily after a conditional \Pull, thus when we reach 
%$r~x$ we necessarily have $\sigma(x)=(V,I)$ or 
%$(V,V)$ (by a trivial case analysis of the possible validity status of $x$ before the 
%\Pull). 
%Other cases are similar.
%\end{proof}


\subsection{Declaring access modes and adding an abstraction layer}
The calculus defined above only considers simple memory locations and directly manipulates 
them.
But VectorPU and  similar libraries manipulate structures 
representing the memory. For example, VectorPU vectors act as an
 abstract representation of a set of memory 
locations. In this section, we add a declaration and abstraction layer to the calculus to 
represent the access mode declarations that will trigger data transfers according to the 
consistency mechanism. 
This abstraction layer is also a necessary first step to the modelling of array 
structures that we will present in Section~\ref{sec-arrays}. Indeed, in array structures, 
the 
validity status of the array is abstracted away by a single validity status pair. Then 
%approximations inspired from classical abstract interpretation techniques~\cite{Cous77} 
a dynamic abstraction of the consistency status of the memory can be used.
%can be used except that the abstract view is computed dynamically but must stay 
%consistent with the validity status of the real memory.
More technically, the abstraction and declaration layer relies on two principles:
\begin{itemize}
\item Each variable $x$ has an abstract variable $\abs x$ that represents it. In this 
section there is 
a single variable for each representative, but when we deal with arrays we will 
have a single representative for the whole array.
\item It is safe to ``forget'' that one memory space holds a valid copy of the data if 
the other memory space has a valid one. In other words, $(V,I)$ (resp. $(I,V)$) is a safe 
abstraction of $(V,V)$ and we denote $(V,I)\leq (V,V)$ (resp. $(I,V)\leq (V,V)$). Of 
course, we have $(X,Y)\leq (X,Y)$ for any $X$ and $Y$.
\end{itemize}

\paragraph{Syntax}
We now define access mode declarations:\\[-3.3ex]
\begin{align*}
\AM&::=R\ \abs x \,|\, W\ \abs x \,|\, RW\ \abs x \,|\, 
\rem{R\ \abs x} \,|\,\rem{W\ \abs x} \,|\,\rem{RW\ \abs x} \,|\, \\
&~ \AM \land \AM' \quad \text{(where variables in $\AM$ and $\AM'$ are disjoint)}
\end{align*}

These access modes declare the kind of access (read $R$, write $W$, or read and/or write 
$RW$) that 
can be performed on the variable $x$ represented by $\abs x$. In a set 
of access mode declarations the same variable cannot appear twice. There exist declared 
access modes for  local accesses and for  the 
remote memory space.

% added "calls to" - one could even use the term "components" from Sect. 2
A program is a sequence of calls to functions or components (i.e., statements accessing 
only real variables) 
each protected by an access 
mode declaration (on abstract variables representing the real variables):
\[\Prog::=\AM_1\{S_1\};\AM_2\{S_2\};\ldots\]
We write  $S\in S'$ if $S$ is one statement inside $S'$ (i.e. $S$ is a sub-term of 
$S'$).

We  define below the semantics of these programs  and specify well-declared program by 
comparing the statements they contain with 
the declared access modes. The semantics relies on the translation of 
the access mode declarations into consistency mechanisms 
with checks and data transfers 
triggered 
before each function 
execution.


\paragraph{Extension of statements to abstract variables}
When evaluating a program, the store contains both real and abstract variables, and the 
existing 
statements have the same effect on the abstract variables as on the real ones. However 
one should notice that 
even if the effect is the same, the meaning of a statement acting on a real variable 
or on its representative is different: in our calculus, the effect on a variable is an 
abstraction of the real effect that involves side effects and data transfers. On the 
contrary, only the validity status of abstract variables is stored by the library: the 
effect triggered by an operation on an abstract variable is exactly what happens when 
VectorPU updates the validity status of its internal structures.

For example, a \Pull\ operation on a real variable consists in transferring data from a 
remote memory space 
to the local one. We abstracted it  by changing the local validity status. A \Pull\  
operation on an abstract variable only changes the validity status, no data transfer has 
to be done because abstract variables only need to be stored in one memory space. 
The validity status is stored in the CPU address space in VectorPU. Comparing the validity 
status of real memory and their representative  allow us to reason 
formally on the correctness of the validity tracking performed by VectorPU.

\begin{figure*}[tb]
\begin{mathpar}
\transl{R~\abs x}=(\IF{\isvalid~\abs x}{\noop}{(\Pull~x;\Pull~\abs x)})

\transl{\rem {R~\abs x}}=(\IF{\isremvalid~x}{\noop}{(\Push~x;\Push~\abs x)})

\transl{RW~\abs x}=(\IF{\isvalid~x}{\noop}{(\Pull~x;\Pull~\abs x)});w~\abs x

\transl{\rem{RW~\abs x}}=(\IF{\isremvalid~x}{\noop}{(\Push~x;\Push~\abs x)});\rem{w~\abs 
x}

\transl{W~\abs x}= w~\abs x

\transl{\rem{W~\abs x}}= \rem{w~\abs x}

\transl{\AM_1\{S_1\};\AM_2\{S_2\};\ldots} = \transl{\AM_1};S_1;\transl{\AM_2};S_2;\ldots
\end{mathpar}
\caption{Semantics of access modes and programs}\label{sem-AM}
\end{figure*}
As no data is accessed by the effects on abstract variables, they cannot create stuck configuration. Consequently, $r~\abs x$ has no effect as it does not 
change the validity 
% Comment: This is only because VectorPU uses the most primitive
% cache coherence protocol, the MI (VI) protocol.
% More elaborated coherence protocols like MSI or MESI introduce additional
% states where also reads trigger state transitions to follow up the
% number of readers (one or larger than one).
status of variables. The statement that should get stuck in case of a read access is the 
read of \emph{the real variable that cannot access a valid data}. 
%Similarly, the interesting 
%effect of read/write ($rw$) operations is entirely represented by the write ($w$) access 
%and $rw$ is not needed over abstract variables.


\paragraph{Semantics}

Figure~\ref{sem-AM} defines the semantics of programs with access modes as a translation 
into the core calculus of Section~\ref{sec-core}. 
This translation
ensures that the validity status is correct and records the effect of the function on the 
abstract variable before running the function call that may 
read and write data (on the real variables).  Similarly to the VectorPU library, the 
protected accesses can be 
considered as macros and the programs can be translated into the 
core syntax.

This encoding corresponds  to the macros as they are implemented in VectorPU.  
It is indeed easy 
to check that VectorPU tracks the effects in 
the same way as our effect system  does in the translation rules. These translation 
rules  perform \Push\ or 
\Pull\ operations in order to ensure that the memory is in a correct validity status for 
the read or write operation to be performed.
When evaluating a program we create a store where the validity status of real and 
abstract variables are $(V,I)$, corresponding to the fact that data is 
initially placed in one memory location; typically, in VectorPU, in the CPU memory space.
%
%
%From this semantics, we state two crucial correctness properties for the library and 
%show 
%their correctness in a semi-formal way.


\subsection{Well-declared Programs and their Properties}
We now define formally what it means for an access mode declaration to be correct, i.e. 
to adequately specify the effect of a function. The principle is that each operation on a 
memory location must be declared on its representative. It is however possible to declare 
more read or $RW$ accesses that what is done in practice, and one can declare a read 
and/or write 
access if only read or write is performed. Additionally, the annotation $W$ denotes an 
\emph{obligation} to write which allows the consistency mechanism to avoid any validity 
check and any transfer before running the function that will overwrite the data. 
To represent this concept, we need a first definition that states that an operation will be performed in all execution paths of a (bigger) statement. This 
definition 
formalises a classical static analysis concept that states that all branches of 
conditionals necessarily execute a given statement. It considers executions that run to 
completion and states that a given statement is necessarily evaluated in this execution.

\begin{definition}[Occur in all execution paths]
We state that a statement\emph{ $S$ occur in all execution paths of $S_0$} if, for any 
correct 
initial store $\sigma_0$, for all full reductions 
$(S_0,\sigma_0)\to(S_1,\sigma_1)\to\ldots\to(\noop,\sigma_n)$, there is an intermediate 
state $(S_i,\sigma_i)$ such that $S_i=S;S''$ for some $S''$.
\end{definition}
Notice that an operation $S$ can only appear in some of the execution paths of $S'$ if  $S\in 
S'$: if $(S_0,\sigma_0)\to^* (S;S',\sigma)$ then $S\in S_0$.


\begin{definition}[Well-declared program]\label{def-WD}
A program $\Prog$ is \emph{well-declared} if for all $\AM\{S\}$ in $\Prog$ we have:
\begin{itemize}
\item $\Push\ x\not\in S$ and $\Pull\ x\not \in S$ (for any $x$),
\item $w\ x\in S \implies (W\ \abs x \in \AM \lor RW\ \abs x \in \AM)$,
\item $r\ x\in S \implies (R\ \abs x \in \AM \lor RW\ \abs x \in \AM)$,
\item $W\ \abs x\!\in\! \AM \!\implies\! w\ x$ occurs in all execution paths of $S$,
\item Plus the same rules for remote operations.
\end{itemize}
\end{definition}
Note that a well-declared program does not perform synchronisation 
operations (\Push\ or \Pull) manually, these operations are only  performed when 
evaluating the 
access mode declarations. Also each variable accessed by a well-declared function has an 
abstract representative in the corresponding declaration block.


A direct consequence of the definition above is that a well-declared program cannot 
access, in the same function, the same
variable in both address spaces. This is in accordance with VectorPU where each function 
is entirely executed either on a CPU or on a GPU, the formalisation is a bit more 
 generic on this aspect. This is expressed by the following property.
\begin{Property}[Localised access]\label{prop-localised}
For a well-declared program containing $\AM\{S\}$, for any $x$, we cannot have $\rem {E\ 
x} \in S$ and $E'\ x \in S$.
\end{Property}

\smallskip

We now state and  prove the two major properties ensured by our formalisation.
The first property ensures that the abstraction is correct relatively to the 
execution. This corresponds to the fact that VectorPU tracks adequately the validity 
status of the 
memory. This is expressed as a theorem that is similar to subject-reduction in type 
systems, it states that if the status of the abstract variables represent correctly the 
validity status of the real variables, then the 
abstraction is also correct after the execution of a  well-declared function.  Let us say 
that  we have a \emph{correct abstraction of the memory state} if for each real memory 
location, the abstract representative of this location has a validity status that is an 
approximation, in the sense of $\leq$, of the validity status of the real memory. The 
theorem below states that the execution of a well-declared function maintains the 
correctness of the memory state 
abstraction. 

\begin{Theorem}[Subject reduction]\label{thm-SR}
Suppose $\AM\{S\}$ is well-declared, we have:
\[\begin{array}{@{}l@{}}
\forall x\in \dom(\sigma).\, \sigma(\abs x)\leq\sigma(x) 
\land
 (\transl{\AM\{S\}},\sigma) \to^* (\noop,\sigma')
\\~~~~~
\implies \forall x\in \dom(\sigma').\, \sigma'(\abs x)\leq\sigma'(x) 
\end{array}\]

This property is extended by a trivial induction to the execution of a well-protected 
program $\Prog$ in an initial store $\sigma_0=(x\mapsto 
(V,I))^{x\in \vars(\Prog)}$.
\end{Theorem}
\begin{proof}
Notice that $\dom(\sigma')=\dom(\sigma)$, and  if $\sigma(x)=(V,I)$ or $\sigma(x)=(I,V)$ 
then $\sigma(x)=\sigma(\abs x)$, else $\sigma(x)=(V,V)$.
We reason on the read and write access that occur in the considered reduction. Each 
variable $x$ is either read or written or not accessed (or read and written). For each 
case we compare the 
status of abstract and local variable, and in particular we consider the status of the 
reduction after executing the synchronisation code $\transl{\AM\{S\}}$ and call 
$\sigma_s$ the corresponding store (note that $\sigma_s(\abs x)=\sigma'(\abs x)$). We 
detail operations on the local 
address space, cases for remote operations are similar:

\noindent$\bullet$ If  $x$ is written, we have:
$(\transl{\AM\{S\}},\sigma)\to^* (w~x;S',\sigma'')  \to^* (\noop,\sigma')$. Whatever the 
initial value of $\sigma(x)$, we have $\sigma'(x)=(V,I)$. Two cases are possible:

\noindent$(1)$ $W~\abs x \in \AM$ then the value cannot be read and we have 
$\sigma_s(\abs x)=(V,I)$.  $ \sigma'(\abs x)=\sigma'(x)$.

\noindent$(2)$ $RW~\abs x \in \AM$ then a data-transfer (\Pull) may occur. Knowing that 
$\sigma(\abs x)\leq\sigma(x)$, by a  
case 
analysis on $\sigma(x)$ and $\sigma(\abs x)$ we have: $\sigma_s(\abs x)=(V,I)$ and 
$\sigma_s(x)=(V,I)$ or $(V,V)$. Whether $x$ is  read or not we have $ \sigma'(\abs 
x)=\sigma'(x)$.

\noindent$\bullet$ If  $x$ is read but not written, its validity status is 
not changed. %Two cases are possible:

\noindent$(1)$ $R~\abs x \in \AM$. By 
a  case 
analysis on $\sigma(x)$ and $\sigma(\abs x)$ we have:
$\sigma_s(x)\!=\!(V,I)$ and  $\sigma_s(\abs x)\!=\!(V,I)$, or $\sigma_s(x)\!=\!(V,V)$ 
and  
$\sigma_s(\abs x)\!=\!(V,I)$ or $(V,V)$.  Reading has no effect on validity status and in 
all 
cases we have $\sigma'(\abs x)~\leq~\sigma'(x)=\sigma_s(x)$.

\noindent$(2)$ $RW~\abs x \in \AM$ then similarly to the case (2) above we have 
$\sigma_s(\abs x)=(V,I)$, additionally $\sigma'(x)=\sigma_s(x)=(V,I)$ or $(V,V)$. In all 
cases $\sigma'(\abs x)\leq\sigma'(x)$.

\noindent$\bullet$ If  $x$ is not accessed but is in the declaration, the 
reasoning is the same as if it was only read. 
Note that the variable cannot be declared 
in write mode, $W~\abs x \in \AM$, by
Definition~\ref{def-WD}.
\end{proof}

Finally, a well-declared program always runs to completion: it never tries to access an 
invalid memory location.

\begin{Theorem}[Progress for well-declared programs]\label{thm-progress}
If a program $\Prog$ is well-declared, then its execution cannot reach a stuck 
configuration.
\end{Theorem}

\begin{proof}
 By 
Property~\ref{prop:stuck}, 
it is sufficient to prove that 
unification on the validity status is always possible. 
We consider a reduction  $(\transl{\AM\{S\}},\sigma) \to^* (S,\sigma_s) \to^* \ldots$ 
similarly  to the proof above.

By definition of well-declared 
programs and because of the signature of effects ($w~x$ cannot be stuck), only two 
cases have to be analysed for the local operations (plus two similar cases for remote 
statements):
 \begin{itemize}
\item \Pull\ operations (on $x$ and $\abs x$) in the translation of $R~\abs x$ or 
$RW~\abs x$. Unification 
requires that $\sigma(x)=(X,V)$ and  $\sigma(\abs x)=(Y,V)$.
\item $r~x$ operation in the evaluation of $S$. Unification 
requires that $\sigma'(x)=(V,X)$ where $\sigma'$ is the store in which the read access is 
to be evaluated.
\end{itemize}
Indeed, access mode declarations do not generate reading operations, and by definition 
function statements contain no \Push\ or \Pull.

Concerning the first case, because of Theorem~\ref{thm-SR}, we have $\sigma(\abs x)\leq 
\sigma(x)$, and because of property~\ref{prop:safe} none of them is $(I,I)$. By case 
analysis on the possible values of $\sigma(\abs x)$ and $\sigma(x)$, it is easy to show 
that $\sigma(x)=(X,V)$ and  $\sigma(\abs x)=(Y,V)$ if we reach the two \Pull\ statements 
that perform data transfers before the execution of the function.

Concerning read access, they should be verified by an induction on the reduction steps 
following the state $(S,\sigma_s)$ showing that, for any variable $x$ that is declared $R$ 
or $RW$, in all states we have $\sigma'(x)=(V,X)$. Indeed, by the same analysis as in the 
proof of 
Theorem~\ref{thm-SR} we know that $\sigma'(x)=(V,X)$. Because of 
Property~\ref{prop-localised} no remote operation is possible on $x$ and thus only $r~x$ 
and $w~x$ operations are possible on $x$, both maintain the invariant $\sigma'(x)=(V,X)$ 
for some $X$.
%
%This is done by case analysis of the effect applied in the translated program, and 
%showing that in each case the effect rule can be applied.
%Take the example of \Pull. \Pull\ only appears upon $R~x$ or $RW~x$ and when  
%$\isvalid~x$ is 
%false. In this case $\sigma(x)=(I,V)$ by property~\ref{prop:safe}. Consequently \Pull\ 
%is 
%not stuck.
%Now considering the effect $r$, $r~x$ can only be obtained by translation of $R~x$. 
%Consequently, $r~x$ appears necessarily after a conditional \Pull, thus when we reach 
%$r~x$ we necessarily have $\sigma(x)=(V,I)$ or 
%$(V,V)$ (by a trivial case analysis of the possible validity status of $x$ before the 
%\Pull). 
%Other cases are similar.
\end{proof}

Considering the example above of a variable written on the CPU, and then read on the GPU, 
a well-declared program encoding this behaviour would be  $RW~\abs x\{w~x\};\rem{R~\abs 
x}\{\rem{r~x}\}$. This code automatically generates the \Push\ instruction that prevents 
the program from being stuck.
\subsection{Effects and Access Mode Declarations for Arrays}\label{sec-arrays}
In array structures, the 
validity status of the whole array is abstracted away by a single validity status pair. 
We extend the syntax for arrays as follows, $x[i]$ denotes the indexed access to an 
element of the array. More precisely the new operations on arrays and their elements are 
(we still have the previous operations on non-array and abstract variables):
\[S::= ... \,|\, r~x[i] \,|\, w~x[i] \]

Synchronisation operations (\Push\ and \Pull) exist for arrays but the whole array is 
synchronised, and we write $\Push~x$ and $\Pull~x$ as above.
All the elements of the array are represented by a single abstract variable: $\abs{x}$ 
represents a safe abstraction of the validity status of all $x[i]$. In other words, as soon as one element of the array $x$ is invalid locally (resp. remotely) the validity status of $\abs x$ can only be $(I,V)$ (res. $(V,I)$).

The semantics of access mode declarations and programs is unchanged because 
synchronisation operations and access mode declarations do not concern array elements.
The concept of well-protected programs must be adapted to the case of array structures, 
and more 
precisely to the fact that several  memory locations are represented by a single abstract 
variable.


\begin{definition}[Well-declared program with array access]\label{def:well-declared-array}
A program $\Prog$ is \emph{well-declared} if for all $\AM\{S\}$ in $\Prog$, additionally 
to the rules of Definition~\ref{def-WD}, we have:
\begin{itemize}
\item $w\ x[i]\in S \implies (W\ \abs x \in \AM \lor RW\ \abs x \in \AM)$,
\item $r\ x[i]\in S \implies (R\ \abs x \in \AM \lor RW\ \abs x \in \AM)$,
\item $W\ \abs x\!\in\! \AM \!\implies\! \forall i\!\!\in\!\!\range(x).\, w\ x[i]$ occurs in all execution 
paths of $S$,
\item Plus the same rules for remote operations.
\end{itemize}
Where $\range(x)$ is the set of valid 
 indexes for an array $x$.
\end{definition}

The other properties are expressed similarly and both \emph{subject-reduction}, 
Theorem~\ref{thm-SR}, and \emph{progress}, Theorem~\ref{thm-progress}, are 
still valid. The only change is the ``correct abstraction of the memory state'' criteria 
that becomes $\forall x\in 
\vars(S).\,\forall i\in\range(x).\, \sigma(\abs x)\leq\sigma(x[i])$ instead of $\forall 
i\in\range(x)$ for arrays. The proofs are similar except in the case of $W~x$ 
declarations where the fact that all elements of the array must be written is necessary 
to ensure that no element is in the status $(I,V)$ (which could not be safely represented 
by $(V,I)$) at the end of the function execution. If we focus on the proof of 
Theorem~\ref{thm-SR}, case ``$x$ is written, sub-case (1) we have $\sigma'(\abs x)=(V,I)$ 
which is a safe abstraction because \emph{all elements have been written}, and thus 
$\sigma'(x[i])=(V,I)$ for all $i$. If one element $j$ was not written, we could have had 
$\sigma'(x[i])=(I,V)$ which would invalidate the theorem. Overall, only arguments about correctness of the abstractions need to be adapted, and we can prove the following theorem.
%
%
%\subsection{Towards Array Accesses: Adding an abstraction layer}
%In the simple formalisation provided here, we only consider simple memory locations and 
%consequently there is no need to have an abstract representation of a set of memory 
%locations. We can consider that the manipulation of the VectorPU abstract structure 
%representing the memory and the memory itself as the same entity. While this point of 
%view allowed us to provide a simple and precise enough representation, and to prove the 
%correctness of the general approach, it becomes not close enough to the real 
%implementation if we want to reason on array structures. In array structures, the 
%validity status of the array is abstracted away by a single validity status pair. Then 
%approximations inspired from classical abstract interpretation techniques~\cite{Cous77} 
%can be used except that the abstract view is computed dynamically but must stay 
%consistent with the validity status of the real memory.
\begin{Theorem}\label{thm-correct-array}
If a program using arrays is well-declared according to Definition~\ref{def:well-declared-array} then its execution verifies both the subject-reduction and the progress property.
\end{Theorem}
\subsection{Discussion: Similarities and differences relatively to VectorPU}



Let us compare the formal definition of the coherence protocol, Figure~\ref{sem-AM} (valid for simple memory locations or arrays), with the VectorPU implementation of the protocol for simple arrays, Figure~\ref{fig:vectorpucoherence}. Except from the order of operations and minor changes, the structure is the same and the form of operations is similar. The main difference is that there is no view of abstract vs. concrete variables, however, if we consider that transfer operations on abstract variables have no effect, and that validity status of concrete variables can be abstracted away, the code is  the same as the formalisation. 

Taking a more global point of view, no verification is performed by the VectorPU framework and thus, the property of ``well-declared programs'' is not checked currently by the framework. Such a check could either be done by a static analysis (involving some approximations, meaning  some correct programs could be rejected), or by runtime verification checking that each function performs exactly the required access. The first solution would require more development and might be tricky, while the second one is not acceptable considering the target application domain because of the overhead involved by the dynamic checks. In the current state of the library, the property ``well-declared programs''  must be ensured by the programmer. The implementation of VectorPU relies on the hypothesis that the function declarations are correct, because of this the current formalisation is a significant step forward as it allows us to express precisely what assumption is made by the library on the programmer's code.

Finally concerning expressiveness, the ``well-declared programs'' definition is a bit more flexible than what VectorPU targets at the moment because it is safe to declare in our framework functions that access some variables on the CPU and others on the GPU, and this is not planned in VectorPU, again due to the supported usage scenario.

These differences highlight interesting improvement directions for VectorPU while we can still consider that the current article  is a faithful formalisation of the library. In the next section, we investigate a feature that is not yet supported by VectorPU, but exists in SkePU. We can somehow consider the theoretical results below as a specification of a future extension of the library.

\section{Overlapping arrays}\label{sec:overlap-array}

In the preceding section we supposed that arrays were well-separated. The preceding abstraction would also be valid for an array that would be split into disjoint entities (such as
VectorPU \texttt{pvector}s) and always used either as the disjoint sub-arrays or as the whole. In this section we extend the framework to take into account overlapping arrays. Here we still consider single dimension arrays for simplicity but multi-dimensional arrays could easily be taken into account.

\subsection{Context and Objectives}

In VectorPU, the first \texttt{pvector} on a vector passed as argument for access on device will (over)allocate space for the entire \texttt{vector}, all subsequent \texttt{pvector} accesses to the same \texttt{vector} can skip the allocation.
 Consequently, two successive data transfers of the same memory location will be written to the same memory location, even if the two initial locations are accessed through different (overlapping) \texttt{pvector}s.
Consequently, on the formal side, if several push or pull are performed on overlapping memory locations, the transferred memory has the same overlaps as the source. This is very important to ensure that no two copies of the same array element will coexist in the same location.

\paragraph{Problem statement}
Overlapping arrays raises several difficulties making the approach currently adopted in VectorPU not adapted. Indeed a single write operation can change the validity status of a cell that belongs to several arrays. Consequently, several access annotations may have to be written for a single operation. As a consequence, some annotations may never be correct: a function that is declared to write all the elements of an array $x$ will necessarily write some elements of the arrays that have an overlap with $x$, these overlapping arrays should thus be annotated and transmitted, or another coherence protocol should be used.

\paragraph{Approach}
In this work we take the decision not to change the coherence protocol of VectorPU and instead work at the access mode declaration level to ensure the consistency of overlapping arrays.

To take into account overlapping arrays in VectorPU, two approaches could be envisioned. A naive solution consists in applying the results for non-overlapping arrays. Indeed, Definition~\ref{def:well-declared-array} of well-declared programs with array accesses is still valid. However,  the programmer now has to annotate more variables because each array access operation may involve several arrays. Due to the array overlap, the programmer should now know all the arrays that are impacted by a function execution, including the overlapping arrays that are not passed as parameter, and the library should be extended to pass them as ``artificial'' parameters. To be more explicit, Definition~\ref{def:well-declared-array} should be extended as follows (with a symmetrical rule for remote writing):

\begin{quote}
For every $y$ overlapping $x$, if $i$ is in the range in common between $x$ and $y$, we have $w\ x[i]\in S \implies (W\ \abs y \in \AM \lor RW\ \abs y \in \AM)$.
\end{quote}

First note that no additional rule is necessary for read operations. Indeed, read operations use the validity status but do not modify it, consequently, read operations do not modify the validity status of other arrays and have no consequence on overlapping arrays.
Note also that the above rule restricts a bit the expressible effects as, for example. $W\ \abs x ; \rem{R}\ \abs y$ cannot be valid if $x$ and $y$ overlap.


\medskip

Though very precise, this approach does not seem realistic and we additionally develop an inference mechanism for access mode declarations in presence of overlapping arrays.
The objective  is to infer the correct annotations on  variables that are not passed as parameters. Knowing the access modes for the function parameters, we infer what operation must be done on other intersecting arrays to ensure the coherency of the system, and we express these additional operations as implicit generated annotations. These additional access mode declarations are inferred, the semantics of these added declarations will result in additional data transfers and validation/invalidation operations that make the program correct. This is less precise as it takes a pessimistic approach on the operations performed by the declared arrays. For example, for any array that is declared $RW$ we will suppose that all the array elements may be read and written, but the approach is safe and mostly automatic. This is presented in Section~\ref{sec:infer-overlap} below.
%
%Another way to see the next two subsection is that the first one extends the well-declared definition for overlapping arrays, and the second one defines an inference mechanism that ensures that a function is well-declared in presence of overlapping arrays, provided the declaration is correct not taking overlapping arrays into consideration, i.e. in the sense of Definition~\ref{def:well-declared-array}.



\subsection{Access mode inference for overlapping arrays}\label{sec:infer-overlap}
Starting from a given set of access mode annotations, we want to infer other access modes that are consequences of the overlaps and the existing annotations.
Because we are only aware of an approximation of the effects (for each array variable, effect is abstracted by a single global effect), the inferred accesses will be approximate but can be a safe over-approximation of the effect of the function.
 Without knowing the real accesses performed by the function, we deduce from the declared access modes, a set of additional ``artificial accesses''.

We consider $\AM$ the set of all access modes declared for a given function and extend it so that the function satisfies the well-declared program requirement even with overlapping arrays.

\begin{table}[!tb]
\begin{mathpar}
\inferrule
{E\ \abs x \in\AM}
{E\ \abs x \in\Overlap \AM}

\inferrule
{\rem{E\ \abs x} \in\AM}
{\rem{E\ \abs x} \in\Overlap \AM}


\inferrule
{W\ \abs x \in\AM \\ W\ \abs y \not\in\AM \\ x \text{ and } y \text{ overlap}}
{RW\  \abs y \in\Overlap \AM}

\inferrule
{RW\ \abs x \in\AM \\ x \text{ and } y \text{ overlap}}
{RW\  \abs y \in\Overlap \AM}

\inferrule
{\rem{W\ \abs x} \in\AM \\ \rem{W\ \abs y} \not\in\AM \\ x \text{ and } y \text{ overlap}}
{\rem{RW\  \abs y} \in\Overlap \AM}

\inferrule
{\rem{RW\ \abs x} \in\AM \\ x \text{ and } y \text{ overlap}}
{\rem{RW\  \abs y} \in\Overlap \AM}

\end{mathpar}
\caption{Extension of access mode annotations to deal with overlapping arrays}\label{Overlap-ext-tab}
\end{table}
To understand the principle of the approach, consider the case where 
$W\ \abs x\in \AM$ then $ \forall i\in\range(x). w\ x[i]$ occurs in all execution 
paths of $S$, and thus for all arrays $y$ overlapping $x$  we must have $(W\ \abs y \in \AM \lor RW\ \abs y \in \AM)$. If we have no additional information we will ensure that $RW\ \abs y$ is also in the set of access mode declarations, which is always safe.


\begin{definition}[Extension of access mode declarations for overlapping arrays]\label{def-overlap-annotation}
Consider a set $\AM$ of access mode declarations.
 The extension for overlapping arrays of $\AM$  is the smallest set $\Overlap \AM$ defined by the  rules in Table~\ref{Overlap-ext-tab}.


\end{definition}

The following theorem states that if the set of function parameters is extended according to the preceding extension, then memory consistency is ensured. Note that it means that a set of artificial parameters are to be added to some functions, in the sense that  data-transfers and validity status modifications  have to be performed on vectors that are not among the original parameters of the function.

\begin{Theorem}\label{thm-correct-array-overlap}
Consider a program that is well-declared according to Definition~\ref{def:well-declared-array}, not taking into account  overlapping arrays. Suppose its access mode declarations are extended according to Definition~\ref{def-overlap-annotation} then the execution of the obtained program verifies both subject reduction and progress, even in presence of overlapping arrays.
\end{Theorem}

\begin{proof}[Proof sketch]
The principle of the proof is to prove that, provided a set $\AM$ correctly declares the accesses performed syntactically by a function on its parameters, the set $\Overlap \AM$ is a correct approximation of the accesses performed by the function on all the arrays of the program, i.e. the parameter arrays and the arrays that overlap the parameter arrays. Then Theorem~\ref{thm-correct-array} will be  sufficient to conclude.

Trivially, the first two rules of Table~\ref{Overlap-ext-tab} are sufficient to conclude about normal function parameters. We now need to ensure that operations on overlapping arrays are well-declared. We focus on non-remote operations and prove that (indirect) operations on overlapping arrays are well-declared, according to Definition~\ref{def:well-declared-array} modified by the additional rule introduced in above: 
\begin{quote}
For every $y$ overlapping $x$, if $i$ is in the range in common between $x$ and $y$, we have $w\ x[i]\in S \implies (W\ \abs y \in \AM \lor RW\ \abs y \in \AM)$.
\end{quote}
By a simple case analysis on the possible annotations and the possible operations performed on the arrays, we deduce that the access modes added by the four last rules of Table~\ref{Overlap-ext-tab} are sufficient to ensure this rule and the symmetrical one for remote writing.
\end{proof}



\subsection{Towards an implementation in VectorPU}


% Tell what is needed
To implement the proposed mechanism that ensure the safety of overlapping vector accesses
in a function call $f$, we need to add to VectorPU the two following components:
\begin{itemize}
    \item A representation $r_v$ that allows to retrieve, for any given \texttt{pvector} $pv$
             of a \texttt{vector} $v$, the set of all other valid \texttt{pvector}s of $v$ 
             that overlap with $pv$.
             $r_v$ is initialized as empty when declaring a new \texttt{vector} $v$,
             queried and/or updated at \texttt{pvector} creations, deletions, and at calls,
             and is removed when $v$ is deallocated.
          %At the vector declaration phase of a program,
          %the representation is constructed by overloading the vector constructor.
          %When a vector finishes its life cycle,
          %it is erased from the representation.
    \item A mechanism which intercepts the function call $f$ and,
          for every vector operand (\texttt{vector} or \texttt{pvector}) $pv$ 
          accessed as W or RW in $f$,
          looks up in the corresponding representation $r_v$ all  
          \texttt{pvector}s overlapping with $pv$. %for every vector $v$ in the argument list
          %of the function call $f$. Then we perform the conservative coherence actions on those vectors\TODO{This should refer back to the technique in Theorem 4, i.e., automatically rewrite the given access mode R or W of an overlapping new operand pvector into RW, to be safe.} retrieved by the look-up.
          For other arguments in $f$ that overlap with $pv$ and have access mode
          R or W, their access mode
          is updated to RW, as proposed in Table~\ref{Overlap-ext-tab}.
          For any other existing \texttt{pvector}s $w$ of $v$ not accessed in $f$
          (but possibly in earlier and/or later calls) 
          that do overlap with $pv$, 
          we append shadow arguments $RW(w)$ to $f$ as suggested by Table~\ref{Overlap-ext-tab}.
          Finally, the intercept mechanism performs, as before, the resulting coherence
          actions (data transfers, status updates) and delivers the call.
          For intercepting the function call, the function call operator is overloaded.
          %\TODO{Can such overloading be done automatically, without programmer assistance? - C.}
\end{itemize}

As a simple example, let us consider the following set of \texttt{pvector}s and call sequence:

\begin{verbatim}
v = new vector(10, ...);
pv1 = new pvector( v, [2:5] );
pv2 = new pvector( v, [4:8] );
pv3 = new pvector( v, [7:9] );
pv4 = new pvector( v, [2:3] );
...
f1( ... R(pv1), R(pv2), ... );
f2( ... W(pv3) ... );  
f3( ... RW(pv4), R(pv2), ... );
\end{verbatim}

Intercepting the function calls, we maintain $r_v$ and update the calls as 
described above.
For the call to \verb.f2., we infer from W(pv3) and the overlap of \verb.pv3. with \verb.pv2.
by Table~\ref{Overlap-ext-tab} that the access mode of \verb.pv2. (not accessed in \verb.f2.) must be 
upgraded to RW, which we do by conceptually appending \verb.RW(pv2).
as a shadow argument to \verb.f2.. The call to \verb.f2. is thus
conceptually rewritten\footnote{As all overlapping
\texttt{pvectors} had been identified before the rewriting, the rule needs
not be applied recursively to the appended shadow arguments, here \texttt{RW(pv2)}.} into

\begin{verbatim}
f2( ... W(pv3) ... , RW(pv2) ); 
\end{verbatim}

\noindent
hence we make sure that the access to \verb.pv2. in the subsequent
call to \verb.f3. will be handled correctly. 
%This leads to invalidation of \verb.pv2., i.e., removal from $r_v$,
%before delivering the actual call to \verb.f2..

%Analogously, we infer at call \verb.f3. that \verb.pv1. must be
%invalidated due to its overlap with argument \verb.pv4..
 
% Summarise the requirement for data structure
%For the first component, we can see that a graph representation is needed
%\TODO{Why a graph? A simple ordered list (or interval tree, for optimization) of valid device copies (pvectors) should suffice? SkePU uses an ordinary list (C++ STL map) for this purpose.}.
%Given a simple three-vector set, one can easily construct a scenario
%that all three vectors overlap with each other in memory address ranges.\TODO{Hard to follow. %If it is important, could you please show some example scenario?}
%For the second component, we notice that the insertion, deletion and the look-up operators is required.
%To summarise, we need a data structure that can represent a graph structure and support efficient insertion, deletion
%and the look-up operators.

% given the requirement, select a data structure that fulfil the requirement
%We can choose dense 2-dimensional matrix to represent the graph.\TODO{Lu, please check if a graph is really necessary here. Also, I do not see how a fixed-size dense matrix could be used here, as valid copies are added and removed dynamically as GPU calls are executed. Or do I overlook something? - C.}
%We need to know the size of 2-dimensional matrix in advance.
%We assume the number of vectors in a program is reasonably small
%so that a dense matrix will not take noticeably large amount of memory
%and we can set in advance a reasonable number for the size,
%or rely on the static analysis of a compiler to estimate a conservative size value.
%The insertion of an overlapping array pair requires to write a 1 in a cell of the dense matrix,
%thus only takes $O(1)$ in time, the deletion operator requires to write a 0 in a cell
%thus also takes $O(1)$. The look-up operator only requires to return a row of the matrix,
%thus also takes $O(1)$. Since we only need a binary number in a cell of the matrix,
%thus the memory size of the dense matrix representation can be further reduced.
%To summarise, the 2D dense matrix representation allows to represent the graph
%and support the required operators efficiently.

It remains to select an appropriate data structure for $r_v$ that allows for
efficient dynamic insertion and removal of \texttt{pvector}s of a vector $v$, 
i.e., index intervals, and efficient lookup of all pvectors that overlap
with a given query interval.
For very small numbers of \texttt{pvector}s of a vector $v$, a simple unordered
list of \texttt{pvector}s is sufficient; this is used e.g.\ in the
smart-container coherence management in SkePU \cite{Dastgeer-IJPP15}. 
For scaling up to larger numbers of \texttt{pvector}s,
a \emph{segment tree} \cite[Sec.~10.3]{Overmars} % : Segment Trees, pp.~224–227
could be used. A segment tree storing $n$ intervals
can be updated dynamically (insertion, removal) in time  $O(\log n)$ and
can retrieve the set of all $k$ intervals overlapping with a query interval
in time $O(k+\log n)$; the space requirements is $O(n\log n)$.

% 180916 CK: I disabled the SkePU section for now,
% as we now have the better VectorPU pvector use case instead.
% For the comparison to SkePU we can write something shorter,
% if at all (just explain why it is more difficult,
% see also the corresponding added remarks in the Conclusion).
%
%\input{skepu.tex}


% -----------------------------------------------------------
\section{A few related works}\label{sec:RW} 
Most of the verification works related to memory consistency focus on coherence 
protocols 
and/or 
weak memory models~\cite{pong1997verification}. 
Among them, one could cite~\cite{Gerth1999}, a formal specification of a caching 
algorithm, and its verification in TLA~\cite{Ladkin1999}. These works shows the 
difficulty to reason on memory coherency, but also that specifications in these models 
should rely on a few simple instructions on the type of memory accessed, a bit similarly 
to this proposal.
Coherence protocols have also been verified using CCS specifications~\cite{Barrio01}. 
These various works are quite different from the approach presented in this paper because 
we rely here on a declarative approach for memory accesses: the programmer declares the 
kind of memory accesses performed by a component, and the consistency mechanism ensures 
that each component accesses a valid memory space.

More recently, and adopting a more language oriented approach, Crary and 
Sullivan~\cite{Crary:POPL:2015} designed 
a calculus for expressing ordering of memory accesses in weak memory models, however we 
are interested here in a much simpler problem where memory access is  sequential 
and clearly identified. 
Even an  extension of this work for parallel processes would  result in a 
simpler model than the ones that exist for weak memory models because of the explicit 
consistency points introduced in the execution by the start/end of each function. 

The closest work to ours is probably~\cite{BJPTSAC16} that defines a memory access calculus similar to ours and prove the 
correctness of a generic cache coherence protocol expressed as part of the semantics of 
the calculus. Compared to this work, we are interested in explicit statements on memory 
accesses and thus the cache consistency is partially ensured  by the programmer 
annotations, making the approach and the properties proven significantly different. Some 
aspects of the approaches could however been made more similar, e.g. by extending our 
work to more than two address spaces or adopting a different  syntax. However our problem 
and formalisation are quite simpler, and we 
believe easier to read, while sufficient for our study.
The same authors also designed a formal model written in Maude~\cite{BJPTMaude16} to 
better understand the possible 
optimisations and the impact of the memory organisation on performance in the context of 
cache coherent multicore architectures. This could be an interesting starting point for 
future works, especially if we 
extend our work to better model the performance aspects of VectorPU and want to reason 
formally on the improved performance obtained by the library. Also from the same authors~\cite{Bijo2017} extends the results described above with parallel spawned task and could be a source of inspiration  to extend our work towards parallel function execution.

\section{Conclusion and future works}\label{sec:conclusion}
In this article we provided a formal approach to verify the consistency of the memory 
accesses in heterogeneous computer systems made of two memory spaces. We formalise the 
operations of memory accesses and memory synchronisation between the two memory spaces 
and prove that a program adequately annotated with informations on the memory accesses 
always access valid memory spaces and tracks correctly which of the memory space contains 
the up-to-date data.

The practical result is that we can verify the coherency mechanism used by the VectorPU 
library and ensure that, additionally to the significant performance benefits of the 
approach, the VectorPU mechanisms is correct and ensures the consistency of the memory 
accesses.

We also extended our model for studying the effect of
operations made on overlapping arrays.
The current implementation of VectorPU supposes that the 
(\texttt{pvector}) array operands always
represent disjoint memory locations, it does not 
take into account overlapping arrays. 
Based on the solution developed in our model,
we described an extension of the VectorPU library that could 
deal safely with overlapping array accesses by
overlapping \texttt{pvector} arguments.

We envision several extensions to this work.
% This old text from the conference was replaced
% by the previous new paragraph:
% ,  the most promising is the study 
% of  the operations made on overlapping arrays. 
% The current implementation of VectorPU supposes that the annotated memory 
% accesses deal with disjoint memory locations, it does not 
% take into account overlapping arrays. Designing an extension of the library that could 
% deal safely with overlapping array is one of the future direction we would like to 
% pursue. 
%Additionally, 
The current article only deals with two memory spaces; the extension to many 
memory spaces (as supported e.g.\ in SkePU)
seems relatively simple but  the mechanism dealing 
with memory transfers between several memory locations becomes a bit more complex; its 
formalisation should be similar.

Moreover, we are interested in the application of our approach to the 
verification of other frameworks. 
Indeed VectorPU uses the most primitive
cache coherence protocol, the VI-protocol.
 More elaborated coherence protocols like MSI or MESI 
 (as used e.g.\ in SkePU \cite{Dastgeer-IJPP15}) 
 introduce additional states where the
 number of readers has to be tracked for example. 
 Also, SkePU uses a more space-efficient management of
 partial vector accesses, the
 coherence protocol itself involves explcitly intersection tests with
 existing copies. 
 Verifying such framework would require 
 a modification of our abstract state representation and a modification of the access 
 mode translational semantics.
%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
\section*{References}
\bibliographystyle{elsarticle-num} 
\bibliography{bibliography}



\end{document}
\endinput
%%
%% End of file `elsarticle-template-num.tex'.
